The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2016-02-14 script/en/chapters/numbertheory.tex
First version of CT Book ed 12. Most important point is the newly includes bitciphers chapter. Further things: Tried to reduce latex warnings amap. Changed wording from Sage to SageMath where visible...

It was important to me to make verifiable statements about the security of the
RSA algorithm, and to add Sage code for as much as possible examples.

@@

\noindent The following three examples\footnote{%
  See chapter~\ref{nt:AppArith1} ``\nameref{nt:AppArith1}''
  for the source code to compute the tables using Sage.}
illustrate the properties of multiplicative inverses (here only the lines
for the factors $5$ und $6$ are listed; not the complete multiplication table).

@@

and the squares (mod $103$) can be calculated beforehand
\begin{eqnarray*}
87^2 & \equiv & 50 {\rm ~(mod~}103),\\
87^4 & \equiv & 50^2 \equiv 28 {\rm ~(mod~}103), \\
87^8 & \equiv & 28^2 \equiv 63 {\rm ~(mod~}103), \\
87^{16} & \equiv & 63^2 \equiv 55 {\rm ~(mod~}103),\\
87^{32} & \equiv & 55^2 \equiv 38 {\rm ~(mod~}103).
\end{eqnarray*}
We have\footnote{%
  See chapter~\ref{nt:AppArith2} ``\nameref{nt:AppArith2}''
  for source code implementing the square and multiply method in Sage,
  which can be used to reproduce the calculations above.
}:
\label{SrcArith2}
\begin{eqnarray*}
87^{43} & \equiv & 87^{32+8+2+1}{\rm ~(mod~}103) \nonumber \\
        & \equiv & 87^{32} * 87^8 * 87^2 * 87 {\rm ~(mod~}103) \nonumber \\ 
    & \equiv & 38 * 63 * 50 * 87 \equiv 85 {\rm ~(mod~}103). \nonumber
\end{eqnarray*}
\end{example}

@@

The general case, where $n$ can be any integer (the sample here had n fixed to
$26$), is considered in chapter \ref{l:NumberTheory_Sage_Number-of-RSA-keys}.
There is a Sage program, calculating the number of all $d$. For all $e$ that
are relatively prime to $J(n)$ we can calculate $d$ as follows using the
Euler-Fermat theorem:

@@

\newpage
\begin{example}{ 1:}\\
The following table~\ref{expmod11}\footnote{%
  The Sage sample~\ref{nt_Sage-code_MultOrder_expmod11} contains the source
  code to generate table~\ref{expmod11}.
  See chapter~\ref{nt:AppArith3a1} ``\nameref{nt:AppArith3a1}''.} 
shows the values $a^i$ mod $11$ for the exponents $i = 1, 2, \cdots, 10$, and
for the bases $a =  1, 2, \cdots, 10$ as well as the resulting value
$ord_{11}(a)$ for each $a$.
\end{example}\\

@@

\begin{definition}\label{def-zth-primitiveroot}
If $a$ and $m$ are co-prime\index{Number!co-prime} and if $ord_m(a) = J(m)$
(i.e. a has maximum order), then we say that $a$ is a {\bf primitive root}
of $m$.\footnote{%
  In chapter~\ref{nt:AppArith3a2} ``\nameref{nt:AppArith3a2}'' there are
  Sage programs to calculate primitive roots.%%\label{primitive-roots-with-sage}
}
\end{definition}

@@

\newpage
\begin{example}{ 2:}\\
The following table~\ref{expmod45}\footnote{%
  The Sage sample~\ref{nt_Sage-code_MultOrder_expmod45} contains the source
  code to generate table~\ref{expmod45}.
  See chapter~\ref{nt:AppArith3b} ``\nameref{nt:AppArith3b}''.}
shows the values $a^i$ mod $45$ for the exponents $i = 1, 2, \cdots, 12$ and
for the bases $a =  1, 2, \cdots, 12$ as well as the resulting value
$ord_{45}(a)$ for each $a$.
\end{example}

@@

\newpage
\begin{example}{ 4:}\\
The following table~\ref{expmod46}\footnote{%
  The Sage sample~\ref{nt_Sage-code_MultOrder_expmod46} contains the source
  code to generate table~\ref{expmod46}.
  See chapter~\ref{nt:AppArith3c} ``\nameref{nt:AppArith3c}''.} 
answers the question as to whether the number $a=7$ is a primitive root of $46$.

@@

\noindent The following tables~\ref{expmod14} and~\ref{expmod22}\footnote{%
  See chapter~\ref{nt:AppArith3d}, ``\nameref{nt:AppArith3d}''
  for the source code to generate the tables~~\ref{expmod14} und~\ref{expmod22}
  using Sage.
} serve as samples to introduce cycle lengths\index{cycle length} -- this is a
topic which goes beyond the multiplicative order.

@@

\noindent The maximum cycle length $l_{max}$ is $J(n)$.\\
For the following tables~\ref{expmod14} and~\ref{expmod22} $J(n)$ is
(according to theorem~\ref{J_of_n}):\\
\indent  - $J(14) = J(2*7) = 1*6 = 6$.\\
\indent  - $J(22) = J(2*11) = 1*10 = 10$.

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
%\clearpage
\newpage
%\hypertarget{Chapter_ElementaryNT_10}{}
\hypertarget{RSABeweis}{}
\section{Proof of the RSA procedure with Euler-Fermat}
\index{RSA}\label{rsabeweis}
Using the Euler-Fermat theorem, we can ``prove'' the
RSA\footnote{The RSA procedure\index{RSA!RSA procedure} is the most common
asymmetric cryptography procedure. Developed in 1978 by Ronald Rivest,
Adi Shamir and Leonard Adleman, it can be used both for signatures and for
encryption.
Cryptographers always associate this procedure with the abbreviation
``\textbf{RSA}'' $-$ the following remark is meant with humor to show that
each letter combination can be used with several meanings: In Britain
the ``Royal Society for the encouragement of Arts, Manufactures \& Commerce''
is commonly known as the ``RSA''.
}
procedure in the group $\mathbb{Z}_n^*$.

@@

\item[{\bf 1.}] Select two distinct random prime numbers\footnote{
Compaq introduced the so-called multi-prime method\index{RSA!multi-prime}
with high marketing effort in 2000.
$n$ was the product of two big and one relative small prime: $n=o*p*q.$ 
With theorem~\ref{J_of_p1..pk} we get:
$J(n)= (o-1)*(p-1)*(q-1). $ This method did not assert itself.\\
One reason probably is, that Compaq claimed a patent \index{Patent} on it. 
Generally there is less understanding in Europe and with the Open Source
Initiative\index{Open Source}, that one can claim patents on algorithms.
But there is really no understanding outside the U.S., that one can get
a patent for a special case (3 factors) of an algorithm (RSA), although
the patent for the general case was almost expired.\\
    JCT\index{JCrypTool} contains the multi-prime RSA method both within
    the {\bf Visuals} menu of the Default Perspective as well as within the
    Algorithm Perspective.
}$^,$\footnote{If the
two primes $p$ and $q$ are equal then $ (m^{e})^{d} \equiv m $ mod $n$ is
not true for all $ m<n$ (although $ e*d \equiv 1$ mod $J(n)$ is fulfilled). 
\begin{example}{:}\\
If $ n=5^2$ then according to theorem~\ref{thm-zth-phinum} it 
is $J(n)=5*4=20, ~ e=3, ~ d=7, ~e*d=21\equiv 1$ mod $J(n).$
\end{example}
But it is $ (5^3)^7\equiv 0 $ mod $ 25$. } $p$ and $q$ and calculate 
$n = p*q$.\footnote{The GISA\index{GISA} (German Information Security Agency) 
recommends, to choose the prime factors $p$ and $q$ almost the same,
but not too close:
$$ 0.5 < |\log_2 (p) - \log_2 (q) | <30. $$
They recommend to generate the primes independently and check that the 
restriction is fulfilled (see \cite{nt:GISA2002}).
}\\
The value $n$ is called the RSA modulus.\footnote{
In CT1\index{CrypTool 1} the RSA modulo is denoted with a capital ``$N$'' .}

@@

Experiences about the ellipsed time of factorization\index{factorization}
with the open source software Pari-GP, Sage, CrypTool 1 and CrypTool 2)
can be found in ``Zeitexperimente zur Faktorisierung''.\footnote{%
R.-H. Schulz und H. Witten: ``Zeitexperimente zur Faktorisierung''. Ein Beitrag
zur Didaktik der Kryptographie'', in LogIn Heft Nr. 166/167 (2010) 113-120,
(available currently only in German)\\
\url{http://bscw.schule.de/pub/bscw.cgi/d864899/Schulz_Witten_Zeit-Experimente.pdf}
}

@@

The following Sage sample shows the very different run times when calculating a gcd and a factorization. The section after this sample will explain the essential part of the method used in \cite{nt:Heninger2012}: Using two trees accelerates the calculation of the gcd pairs a lot.

The Sage sample~\ref{nt_sagesample_Compare-Runtime-gcd-factoring} shows that multiplication of factors, dividing a modul with a known factor, or calculating the gcd is very fast. However, factoring moduli steeply increases with longer moduli. Even the relatively small moduli used in this example show this: The smaller modul (69 decimal digits, 228 bit) took 76 seconds, while the bigger one (72 decimal digits, 239 bit) took almost 217 seconds.

@@

%TODO: E: Wohl durch das verschobene Sage-Beispiel wirkt das \newpage vor dem ff. Kapitel nicht!

@@

\newpage
\begin{remark}{:}\\
In this example using such small numbers, it is easily possible to calculate the discrete
logarithms, but with large numbers the discrete logarithm\index{Discrete logarithm}
problem\footnote{%
You can use Sage to determine the discrete logarithm $x$ that solves the
equation $11^x \equiv 49 \pmod{347}$ (here for Alice)\index{Sage}:
\texttt{discrete\_log(mod(49, 347), mod(11, 347))}. The returned value
is $67$.\\
Such number theoretic tasks can also be solved using other
tools like PariGP\index{Pari-GP}, LiDIA\index{LiDIA}, BC\index{BC} or
Mathematica\index{Mathematica}
(see the list of web sites in the appendix at the end of this chapter):\\
- Pari-GP: \texttt{znlog(Mod(49,347),Mod(11,347))}.\\
- LiDIA:   \texttt{dl(11,49,347)}.\\
- Mathematica: The general Solve function delivers the {em tdep message} ``The equations
  appear to involve the variables to be solved for in an essentially non-algebraic way''.\\
- Mathematica: {\tt MultiplicativeOrder[11, 347, 49]}.\\
All deliver the result $67$.
}${}^,$\footnote{%
Why have the functions delivered the value $67$ for the discrete
logarithm\index{Discrete logarithm} of Alice rather than $240$ which Alice
selected as exponent $a$?\\
The discrete logarithm is
the smallest natural exponent that solves the equation 
$11^x \equiv 49 \pmod{347}$. Both $x = 67$ and $x = 240$ (the number
selected in the example) satisfy the equation and can therefore be
used to calculate the session key: 
$285^{240} \equiv 285^{67} \equiv 268 \pmod{347}$.
If Alice and Bob
had selected a primitive root\index{Primitive root} modulo $p$ as base $g$, then for every
remainder from the set $\{1, 2, \dots, p-1 \}$ there is exactly one
exponent from the set $\{0, 1, \dots, p-2 \}$. \\
\indent As an aside, there are $172$ different primitive roots modulo $347$,
$32$ of which are prime (not necessary). Since the number $11$
selected for $g$ in the example is not a primitive root\index{Primitive root} of $347$, the
remainders do not take all values from the set $\{1, 2, \dots, 346 \}$. 
Thus, for a particular remainder there may be more than one exponent
or even no exponent at all in the set $\{0, 1, \dots, 345 \}$ that
satisfies the equation.\\
With the relevant Sage\index{Sage} commands you find:\\
\texttt{is\_prime(347)=True}, \texttt{euler\_phi(347)=346}, \texttt{gcd(11,347)=1} and 
\texttt{multiplicative\_order(mod(11, 347))=173}.

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\hypertarget{Chapter_ElementaryNT_12}{}
\section[The RSA procedure with actual numbers]{The RSA procedure with actual numbers\footnotemark}
\footnotetext{%
    \index{Sage}%
    \index{Nguyen, Minh Van }%
    Additional material: Minh Van Nguyen, ``Number Theory and the RSA
    Public Key Cryptosystem'', 2009. An introductory tutorial on using
    Sage to study elementary number theory and public key
    cryptography. A didactically very clear article about some basic
    number theory and Sage usage.\\
    \url{http://nguyenminh2.googlepages.com/sage_numtheory-rsa.pdf}.
}
\label{rsaconcrete}\index{RSA}

@@

\label{SrcArith4a}
\noindent All 7 parts are encrypted using: $C \equiv M^{37}{\rm ~(mod~}3713)$:\footnote{%
  See chapter~\ref{nt:AppArith4a} ``\nameref{nt:AppArith4a}''
  for source code to do RSA encryption using Sage.
} \\
{\tt 1404 2932 3536 0001 3284 2280 2235}

@@

\noindent We will work through the example in 2 variants. The steps 1 to 3 are common for both.
\par
\noindent {\bf Key generation (steps 1 to 3)}:
\label{SrcArith4b} \\
{\bf 1.} $p=503,~q=509 \quad (n= 256,027; \; J(n)=(p-1)(q-1)=255,016=2^3*127*251)$.\footnote{%
  See chapter~\ref{nt:AppArith4b} ``\nameref{nt:AppArith4b}''
  for the source code to factorize the number $J(n)$ using Sage.
} \\
{\bf 2.} $e=65,537$ \\
\strut\quad\ ($e$ should\footnote{%
                See footnote~\ref{foot:Selection-of-e} on page
                \pageref{foot:Selection-of-e}.}
                lie between $509$ and $255,015$, and must\footnote{%
       $e$ cannot, therefore, be $2, 127$ or $251$
       ($65,537 = 2^{16}+1$) ($255,016 = 2^{3}*127*251$).\\
       In real life, $J(n)$ is not factorized but rather the Euclidean
       algorithm is used for the selected e to guarantee that
       ${\rm gcd}(e,J(n))=1$.
                 } be relatively prime to $255,016$).\\
{\bf 3.} $d=231,953$ \\
\strut\quad\ (since $e \equiv d^{-1}{\rm ~mod~}J(n): ~65,537*231,953 \equiv 15,201,503,761 \equiv 1
{\rm ~(mod~}255,016)$).\footnote{Other possible combinations of $(e,d)$ include: $(3, 170,011)$, $(5, 204,013)$, $(7, 36,431)$.}

@@

\noindent The letters are not combined!\footnote{For secure procedures we need large numbers that assume -- as far as possible -- all 
values up to $n-1$. If the possible value set for the numbers in the message is too small, even large prime numbers 
cannot make the procedure secure.  An ASCII character is represented by $8$ bits. If we want larger values we must 
combine several numbers. Two characters need $16$ bits, whereby the maximum value that can be represented is $65536$. 
The modulus $n$ must then be greater than $2^{16} = 65536$. This is applied in variant 2.
When the numbers are combined, the leading zeros are kept in binary notation (just as if we were to write all numbers with $3$ 
digits in decimal notation above and were then to obtain the sequence 
{\tt 082 083,  065 032, 119 111,  114 107,  115 033}).}
\label{SrcArith4c}\\
Each character is encrypted using: $C = M^{65,537} {\rm ~(mod~} 256,027)$:\footnote{%
  See chapter~\ref{nt:AppArith4c} ``\nameref{nt:AppArith4c}''
  for the source code for RSA exponentiation using Sage.
} \\
{\tt
\begin{tabular}{lllll}
212984 & 025546 & 104529 & 031692 & 248407 \\
100412 & 054196 & 100184 & 058179 & 227433\\
\end{tabular} }

@@

\label{SrcArith4d}
\noindent Each block is encrypted using: $C \equiv M^{65,537}{\rm ~(mod~}256,027)$:\footnote{%
  See chapter~\ref{nt:AppArith4d} ``\nameref{nt:AppArith4d}''
  for the source code for RSA exponentiation using Sage.
} \\
{\tt 158721 137346 37358 240130 112898}

@@

\label{SrcArith4e}
\noindent Each block is encrypted using: $C \equiv M^{65,537}{\rm ~(mod~}256,027)$:\footnote{%
  See chapter~\ref{nt:AppArith4e} ``\nameref{nt:AppArith4e}''
  for the source code for RSA exponentiation using Sage.
} \\
{\tt 198967 051405 254571 115318 014251}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\hypertarget{NumberTheory_Appendix_E}{}
\section{Appendix: Examples using Sage}
\label{NumberTheory_Appendix_E}{}
\index{Sage}
\index{Sage!Code examples}

@@

\noindent Below you can find Sage source code related to contents of the
chapter~\ref{Chapter_ElementaryNT} (``\nameref{Chapter_ElementaryNT}'').

@@

The multiplication table~\ref{mulmod17} (from page \pageref{SrcArith1a})
for $a \times i \pmod{m}$, where
$m = 17$, $a=5$ and $a=6$, and $i$ ranges over all integers from $0$ to $16$
can be computed using Sage as follows:

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: m = 17; a = 5; b = 6
sage: [mod(a * i, m).lift() for i in xrange(m)]
[0, 5, 10, 15, 3, 8, 13, 1, 6, 11, 16, 4, 9, 14, 2, 7, 12]
sage: [mod(b * i, m).lift() for i in xrange(m)]
[0, 6, 12, 1, 7, 13, 2, 8, 14, 3, 9, 15, 4, 10, 16, 5, 11]
\end{Verbatim}
\caption{Multiplication tables for $a \times i \pmod{m}$ with $m = 17$, $a=5$ and $a=6$}
\end{sagecode}

@@

The fast exponentiation modulo $m$ can be computed using the Sage
function \verb!power_mod()!. The result of this function is an
integer. We can compute the exponentiation in the example in chapter
``\nameref{hohpot}'' on page \pageref{SrcArith2} as follows:

>>>>

It was important to me to make verifiable statements about the security of the
RSA algorithm, and to add SageMath code for as much as possible examples.

@@

\noindent The following three examples\footnote{%
  See chapter~\ref{nt:AppArith1} ``\nameref{nt:AppArith1}''
  for the source code to compute the tables using SageMath.}
illustrate the properties of multiplicative inverses (here only the lines
for the factors $5$ und $6$ are listed; not the complete multiplication table).

@@

and the squares (mod $103$) can be calculated beforehand
\begin{eqnarray*}
87^2 & \equiv & 50 {\rm ~(mod~}103),\\
87^4 & \equiv & 50^2 \equiv 28 {\rm ~(mod~}103), \\
87^8 & \equiv & 28^2 \equiv 63 {\rm ~(mod~}103), \\
87^{16} & \equiv & 63^2 \equiv 55 {\rm ~(mod~}103),\\
87^{32} & \equiv & 55^2 \equiv 38 {\rm ~(mod~}103).
\end{eqnarray*}
We have\footnote{%
  See chapter~\ref{nt:AppArith2} ``\nameref{nt:AppArith2}''
  for source code implementing the square and multiply method in SageMath,
  which can be used to reproduce the calculations above.
}:
\label{SrcArith2}
\begin{eqnarray*}
87^{43} & \equiv & 87^{32+8+2+1}{\rm ~(mod~}103) \nonumber \\
        & \equiv & 87^{32} * 87^8 * 87^2 * 87 {\rm ~(mod~}103) \nonumber \\ 
    & \equiv & 38 * 63 * 50 * 87 \equiv 85 {\rm ~(mod~}103). \nonumber
\end{eqnarray*}
\end{example}

@@

The general case, where $n$ can be any integer (the sample here had n fixed to
$26$), is considered in chapter \ref{l:NumberTheory_Sage_Number-of-RSA-keys}.
There is a SageMath program, calculating the number of all $d$. For all $e$ that
are relatively prime to $J(n)$ we can calculate $d$ as follows using the
Euler-Fermat theorem:

@@

\newpage
\begin{example}{ 1:}\\
The following table~\ref{expmod11}\footnote{%
  The SageMath sample~\ref{nt_Sage-code_MultOrder_expmod11} contains the source
  code to generate table~\ref{expmod11}.
  See chapter~\ref{nt:AppArith3a1} ``\nameref{nt:AppArith3a1}''.} 
shows the values $a^i$ mod $11$ for the exponents $i = 1, 2, \cdots, 10$, and
for the bases $a =  1, 2, \cdots, 10$ as well as the resulting value
$ord_{11}(a)$ for each $a$.
\end{example}\\

@@

\begin{definition}\label{def-zth-primitiveroot}
If $a$ and $m$ are co-prime\index{Number!co-prime} and if $ord_m(a) = J(m)$
(i.e. a has maximum order), then we say that $a$ is a {\bf primitive root}
of $m$.\footnote{%
  In chapter~\ref{nt:AppArith3a2} ``\nameref{nt:AppArith3a2}'' there are
  SageMath programs to calculate primitive roots.%%\label{primitive-roots-with-sage}
}
\end{definition}

@@

\newpage
\begin{example}{ 2:}\\
The following table~\ref{expmod45}\footnote{%
  The SageMath sample~\ref{nt_Sage-code_MultOrder_expmod45} contains the source
  code to generate table~\ref{expmod45}.
  See chapter~\ref{nt:AppArith3b} ``\nameref{nt:AppArith3b}''.}
shows the values $a^i$ mod $45$ for the exponents $i = 1, 2, \cdots, 12$ and
for the bases $a =  1, 2, \cdots, 12$ as well as the resulting value
$ord_{45}(a)$ for each $a$.
\end{example}

@@

\newpage
\begin{example}{ 4:}\\
The following table~\ref{expmod46}\footnote{%
  The SageMath sample~\ref{nt_Sage-code_MultOrder_expmod46} contains the source
  code to generate table~\ref{expmod46}.
  See chapter~\ref{nt:AppArith3c} ``\nameref{nt:AppArith3c}''.} 
answers the question as to whether the number $a=7$ is a primitive root of $46$.

@@

\noindent The following tables~\ref{expmod14} and~\ref{expmod22}\footnote{%
  See chapter~\ref{nt:AppArith3d}, ``\nameref{nt:AppArith3d}''
  for the source code to generate the tables~~\ref{expmod14} und~\ref{expmod22}
  using SageMath.
} serve as samples to introduce cycle lengths\index{cycle length} -- this is a
topic which goes beyond the multiplicative order.

@@

\noindent The maximum cycle length $l_{max}$ is $J(n)$.

\noindent For the following tables~\ref{expmod14} and~\ref{expmod22} $J(n)$ is
(according to theorem~\ref{J_of_n}):\\
\indent  - $J(14) = J(2*7) = 1*6 = 6$.\\
\indent  - $J(22) = J(2*11) = 1*10 = 10$.

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\clearpage
\newpage
%\hypertarget{Chapter_ElementaryNT_10}{}
\hypertarget{RSABeweis}{}
\section{Proof of the RSA procedure with Euler-Fermat}
\index{RSA}\label{rsabeweis}
Using the Euler-Fermat theorem, we can ``prove'' the
RSA\footnote{The RSA procedure\index{RSA!RSA procedure} is the most common
asymmetric cryptography procedure. Developed in 1978 by Ronald Rivest,
Adi Shamir and Leonard Adleman, it can be used both for signatures and for
encryption.
Cryptographers always associate this procedure with the abbreviation
``\textbf{RSA}'' $-$ the following remark is meant with humor to show that
each letter combination can be used with several meanings: In Britain
the ``Royal Society for the encouragement of Arts, Manufactures \& Commerce''
is commonly known as the ``RSA''.
}
procedure in the group $\mathbb{Z}_n^*$.

@@

\item[{\bf 1.}] Select two distinct random prime numbers\footnote{
Compaq introduced the so-called multi-prime method\index{RSA!multi-prime}
with high marketing effort in 2000.
$n$ was the product of two big and one relative small prime: $n=o*p*q.$ 
With theorem~\ref{J_of_p1..pk} we get:
$J(n)= (o-1)*(p-1)*(q-1). $ This method did not assert itself.\\
One reason probably is, that Compaq claimed a patent \index{Patent} on it. 
Generally there is less understanding in Europe and with the Open Source
Initiative\index{Open Source}, that one can claim patents on algorithms.
But there is really no understanding outside the U.S., that one can get
a patent for a special case (3 factors) of an algorithm (RSA), although
the patent for the general case was almost expired.\\
    JCT\index{JCrypTool} contains the multi-prime RSA method both within
    the {\bf Visuals} menu of the Default Perspective as well as within the
    Algorithm Perspective.
}$^,$\footnote{If the
two primes $p$ and $q$ are equal then $ (m^{e})^{d} \equiv m $ mod $n$ is
not true for all $ m<n$ (although $ e*d \equiv 1$ mod $J(n)$ is fulfilled). 
\begin{example}{:}\\
If $ n=5^2$ then according to theorem~\ref{thm-zth-phinum} it 
is $J(n)=5*4=20, ~ e=3, ~ d=7, ~e*d=21\equiv 1$ mod $J(n).$
\end{example}
But it is $ (5^3)^7\equiv 0 $ mod $ 25$. } $p$ and $q$ and calculate 
$n = p*q$.\footnote{The GISA\index{GISA} (German Information Security Agency) 
recommends, to choose the prime factors $p$ and $q$ almost the same,
but not too close:
$$ 0.5 < |\log_2 (p) - \log_2 (q) | <30. $$
They recommend to generate the primes independently and check that the 
restriction is fulfilled (see \cite{nt:GISA2015}).
}\\
The value $n$ is called the RSA modulus.\footnote{
In CT1\index{CrypTool 1} the RSA modulo is denoted with a capital ``$N$'' .}

@@

Experiences about the ellipsed time of factorization\index{factorization}
with the open source software Pari-GP, SageMath, CrypTool 1 and CrypTool 2)
can be found in ``Zeitexperimente zur Faktorisierung''.\footnote{%
R.-H. Schulz und H. Witten: ``Zeitexperimente zur Faktorisierung''. Ein Beitrag
zur Didaktik der Kryptographie'', in LogIn Heft Nr. 166/167 (2010) 113-120,
(available currently only in German)\\
\url{http://bscw.schule.de/pub/bscw.cgi/d864899/Schulz_Witten_Zeit-Experimente.pdf}
}

@@

The following SageMath sample shows the very different run times when calculating a gcd and a factorization. The section after this sample will explain the essential part of the method used in \cite{nt:Heninger2012}: Using two trees accelerates the calculation of the gcd pairs a lot.

The SageMath sample~\ref{nt_sagesample_Compare-Runtime-gcd-factoring} shows that multiplication of factors, dividing a modul with a known factor, or calculating the gcd is very fast. However, factoring moduli steeply increases with longer moduli. Even the relatively small moduli used in this example show this: The smaller modul (69 decimal digits, 228 bit) took 76 seconds, while the bigger one (72 decimal digits, 239 bit) took almost 217 seconds.

@@

%TODO: E: Wohl durch das verschobene SageMath-Beispiel wirkt das \newpage vor dem ff. Kapitel nicht!

@@

\newpage
\begin{remark}{:}\\
In this example using such small numbers, it is easily possible to calculate the discrete
logarithms, but with large numbers the discrete logarithm\index{Discrete logarithm}
problem\footnote{%
You can use SageMath to determine the discrete logarithm $x$ that solves the
equation $11^x \equiv 49 \pmod{347}$ (here for Alice)\index{SageMath}:
\texttt{discrete\_log(mod(49, 347), mod(11, 347))}. The returned value
is $67$.\\
Such number theoretic tasks can also be solved using other
tools like PariGP\index{Pari-GP}, LiDIA\index{LiDIA}, BC\index{BC} or
Mathematica\index{Mathematica}
(see the list of web sites in the appendix at the end of this chapter):\\
- Pari-GP: \texttt{znlog(Mod(49,347),Mod(11,347))}.\\
- LiDIA:   \texttt{dl(11,49,347)}.\\
- Mathematica: The general Solve function delivers the {em tdep message} ``The equations
  appear to involve the variables to be solved for in an essentially non-algebraic way''.\\
- Mathematica: {\tt MultiplicativeOrder[11, 347, 49]}.\\
All deliver the result $67$.
}${}^,$\footnote{%
Why have the functions delivered the value $67$ for the discrete
logarithm\index{Discrete logarithm} of Alice rather than $240$ which Alice
selected as exponent $a$?\\
The discrete logarithm is
the smallest natural exponent that solves the equation 
$11^x \equiv 49 \pmod{347}$. Both $x = 67$ and $x = 240$ (the number
selected in the example) satisfy the equation and can therefore be
used to calculate the session key: 
$285^{240} \equiv 285^{67} \equiv 268 \pmod{347}$.
If Alice and Bob
had selected a primitive root\index{Primitive root} modulo $p$ as base $g$, then for every
remainder from the set $\{1, 2, \dots, p-1 \}$ there is exactly one
exponent from the set $\{0, 1, \dots, p-2 \}$. \\
\indent As an aside, there are $172$ different primitive roots modulo $347$,
$32$ of which are prime (not necessary). Since the number $11$
selected for $g$ in the example is not a primitive root\index{Primitive root} of $347$, the
remainders do not take all values from the set $\{1, 2, \dots, 346 \}$. 
Thus, for a particular remainder there may be more than one exponent
or even no exponent at all in the set $\{0, 1, \dots, 345 \}$ that
satisfies the equation.\\
With the relevant SageMath\index{SageMath} commands you find:\\
\texttt{is\_prime(347)=True}, \texttt{euler\_phi(347)=346}, \texttt{gcd(11,347)=1} and 
\texttt{multiplicative\_order(mod(11, 347))=173}.

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\hypertarget{Chapter_ElementaryNT_12}{}
\section[The RSA procedure with actual numbers]{The RSA procedure with actual numbers\footnotemark}
\footnotetext{%
    \index{SageMath}%
    \index{Nguyen, Minh Van }%
    Additional material: Minh Van Nguyen, ``Number Theory and the RSA
    Public Key Cryptosystem'', 2009. An introductory tutorial on using
    SageMath to study elementary number theory and public key
    cryptography. A didactically very clear article about some basic
    number theory and SageMath usage.%\\
%    \url{http://nguyenminh2.googlepages.com/sage_numtheory-rsa.pdf}.
}
\label{rsaconcrete}\index{RSA}

@@

\label{SrcArith4a}
\noindent All 7 parts are encrypted using: $C \equiv M^{37}{\rm ~(mod~}3713)$:\footnote{%
  See chapter~\ref{nt:AppArith4a} ``\nameref{nt:AppArith4a}''
  for source code to do RSA encryption using SageMath.
} \\
{\tt 1404 2932 3536 0001 3284 2280 2235}

@@

\noindent We will work through the example in 2 variants. The steps 1 to 3 are common for both.
\par
\noindent {\bf Key generation (steps 1 to 3)}:
\label{SrcArith4b} \\
{\bf 1.} $p=503,~q=509 \quad (n= 256,027; \; J(n)=(p-1)(q-1)=255,016=2^3*127*251)$.\footnote{%
  See chapter~\ref{nt:AppArith4b} ``\nameref{nt:AppArith4b}''
  for the source code to factorize the number $J(n)$ using SageMath.
} \\
{\bf 2.} $e=65,537$ \\
\strut\quad\ ($e$ should\footnote{%
                See footnote~\ref{foot:Selection-of-e} on page
                \pageref{foot:Selection-of-e}.}
                lie between $509$ and $255,015$, and must\footnote{%
       $e$ cannot, therefore, be $2, 127$ or $251$
       ($65,537 = 2^{16}+1$) ($255,016 = 2^{3}*127*251$).\\
       In real life, $J(n)$ is not factorized but rather the Euclidean
       algorithm is used for the selected e to guarantee that
       ${\rm gcd}(e,J(n))=1$.
                 } be relatively prime to $255,016$).\\
{\bf 3.} $d=231,953$ \\
\strut\quad\ (since $e \equiv d^{-1}{\rm ~mod~}J(n): ~65,537*231,953 \equiv 15,201,503,761 \equiv 1
{\rm ~(mod~}255,016)$).\footnote{Other possible combinations of $(e,d)$ include: $(3, 170,011)$, $(5, 204,013)$, $(7, 36,431)$.}

@@

\noindent The letters are not combined!\footnote{For secure procedures we need large numbers that assume -- as far as possible -- all 
values up to $n-1$. If the possible value set for the numbers in the message is too small, even large prime numbers 
cannot make the procedure secure.  An ASCII character is represented by $8$ bits. If we want larger values we must 
combine several numbers. Two characters need $16$ bits, whereby the maximum value that can be represented is $65536$. 
The modulus $n$ must then be greater than $2^{16} = 65536$. This is applied in variant 2.
When the numbers are combined, the leading zeros are kept in binary notation (just as if we were to write all numbers with $3$ 
digits in decimal notation above and were then to obtain the sequence 
{\tt 082 083,  065 032, 119 111,  114 107,  115 033}).}
\label{SrcArith4c}\\
Each character is encrypted using: $C = M^{65,537} {\rm ~(mod~} 256,027)$:\footnote{%
  See chapter~\ref{nt:AppArith4c} ``\nameref{nt:AppArith4c}''
  for the source code for RSA exponentiation using SageMath.
} \\
{\tt
\begin{tabular}{lllll}
212984 & 025546 & 104529 & 031692 & 248407 \\
100412 & 054196 & 100184 & 058179 & 227433\\
\end{tabular} }

@@

\label{SrcArith4d}
\noindent Each block is encrypted using: $C \equiv M^{65,537}{\rm ~(mod~}256,027)$:\footnote{%
  See chapter~\ref{nt:AppArith4d} ``\nameref{nt:AppArith4d}''
  for the source code for RSA exponentiation using SageMath.
} \\
{\tt 158721 137346 37358 240130 112898}

@@

\label{SrcArith4e}
\noindent Each block is encrypted using: $C \equiv M^{65,537}{\rm ~(mod~}256,027)$:\footnote{%
  See chapter~\ref{nt:AppArith4e} ``\nameref{nt:AppArith4e}''
  for the source code for RSA exponentiation using SageMath.
} \\
{\tt 198967 051405 254571 115318 014251}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\hypertarget{NumberTheory_Appendix_E}{}
\section{Appendix: Examples using SageMath}
\label{NumberTheory_Appendix_E}{}
\index{SageMath}
\index{SageMath!Code examples}

@@

\noindent Below you can find SageMath source code related to contents of the
chapter~\ref{Chapter_ElementaryNT} (``\nameref{Chapter_ElementaryNT}'').

@@

The multiplication table~\ref{mulmod17} (from page \pageref{SrcArith1a})
for $a \times i \pmod{m}$, where
$m = 17$, $a=5$ and $a=6$, and $i$ ranges over all integers from $0$ to $16$
can be computed using SageMath as follows:

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize]
sage: m = 17; a = 5; b = 6
sage: [mod(a * i, m).lift() for i in xrange(m)]
[0, 5, 10, 15, 3, 8, 13, 1, 6, 11, 16, 4, 9, 14, 2, 7, 12]
sage: [mod(b * i, m).lift() for i in xrange(m)]
[0, 6, 12, 1, 7, 13, 2, 8, 14, 3, 9, 15, 4, 10, 16, 5, 11]
\end{Verbatim}
\caption{Multiplication tables for $a \times i \pmod{m}$ with $m = 17$, $a=5$ and $a=6$}
\end{sagecode}

@@

The fast exponentiation modulo $m$ can be computed using the SageMath
function \verb!power_mod()!. The result of this function is an
integer. We can compute the exponentiation in the example in chapter
``\nameref{hohpot}'' on page \pageref{SrcArith2} as follows:
